The Origins of Congestion and\\Network Assisted End-to-End Congestion Control

### 2020-05-21 paper/paper.tex
Citation Titles didn't have upper-case error fix

\section{Congestion Avoidance}
\label{sec:ca}
As already mentioned, congestion detection algorithms make up an important part in congestion avoidance protocols. \textit{TCP} is such a protocol, based on the transport layer, which adapts its send rate according to available network bandwidth \cite{1209197,jacobson1992tcp}. To fully understand, how TCP's congestion avoidance algorithm operates, one needs to understand how TCP works. TCP lives on the transport layer. It communicates with the application layer via sockets, splits up data into packets (and vice versa), and transmits those packages over the network with the help of the internet protocol, thus TCP is often also referred to as \textit{TCP/IP}. It is the interface between network components and computer programs. Each of those packets have different purposes, but underlay the same structure - the protocol header. The TCP header is split up into multiple segments [Figure \ref{fig:TCP_Header}]:
\begin{figure}
\centering
\begin{bytefield}[bitheight=2.2\baselineskip ,bitwidth=0.635\baselineskip]{32}
	\bitbox{16}{Source Port} & \bitbox{16}{Destination Port} \\
	\bitbox{32}{Sequence Number} \\
	\bitbox{32}{Acknowledgment Number} \\
	\bitbox{4}{Data Offset} & \bitbox{4}{Rsrvd} &
\bitbox{1}{\tiny C\\W\\R} & \bitbox{1}{\tiny E\\C\\E} &
\bitbox{1}{\tiny U\\R\\G} & \bitbox{1}{\tiny A\\C\\K} &
\bitbox{1}{\tiny P\\S\\H} & \bitbox{1}{\tiny R\\S\\T} &
\bitbox{1}{\tiny S\\Y\\N} & \bitbox{1}{\tiny F\\I\\N} &
\bitbox{16}{Window Size} \\
\bitbox{16}{Checksum} & \bitbox{16}{Urgent Pointer} \\
\bitbox{32}{Options} \\
\bitbox{32}{Data}
\end{bytefield}
\caption{Basic structure of the TCP header. It is split up into multiple 32-bit blocks. The data block contains all the application data to be transmitted \cite{ietf-tcpm-rfc793bis-16,jacobson1992tcp,huston2000tcp}.}
\label{fig:TCP_Header}
\end{figure}
\begin{itemize}
\item \textit{Source Port} and \textit{Destination Port} specify the ports on which TCP should communicate with the application layer.

@@

\item TCP can set eight different bit-flags in its header. The most important one is the \textit{ACK} flag (Acknowledgement bit). It is used whenever a packet arrives at its destination and an acknowledgement packet has to be sent back. The \textit{SYN} and \textit{FIN} flags initialize and terminate a connection.

\item The \textit{Window Size} is the maximum amount of data (in bytes) a receiver is willing to receive at any time. It is being negotiated at the initialization of a new connection and can only be negotiated again wile the \textit{SYN} flag is set. It is important to set it appropriately to prevent unnecessary packet loss or delay, as the receiver couldn't process the data in time \cite{jacobson1992tcp}.
\end{itemize}
TCP uses a handshake system. Whenever a client sends a data packet to the server, the server waits until it has received enough data (specified by the window size), and then sends back an acknowledgment package, signaling the successful transmission of data. There are also other handshake operations used, for example for establishing or terminating a connection, however, those have no further importance to the congestion avoidance protocol \cite{1209197,jacobson1992tcp,huston2000tcp,jacobson1995congestion}.

\subsection{Perceiving Congestion}
\label{P_C_marker}
Both partners of a connection are both sender and receiver, as both transmit packets. In the following however, we will simplify this and specify one partner as the sender, the other as the receiver. The sender keeps track of the time it takes from when a package sent until the acknowledgment is received. This is also known as \textit{round-trip time} (RTT). Comparing the current RTT to previously measured RTTs can inform the sender about the current state of congestion. Some variances of TCP make use of this feature in their congestion avoidance algorithms, yet classic TCP does not \cite{huston2000tcp,jacobson1995congestion}. RTT also plays a very important role in the detection of packet loss, which further implies congestion. The sender side keeps track of the RTT while it is being measured and compares it to the \textit{retransmission timeout} (RTO) \cite{jacobson1992tcp}. RTO time is being calculated by estimating the mean and variance of the RTT. Whenever packets are not acknowledged within this RTO interval, the TCP sender regards this packet as lost and will retransmit its data. Therefore, it is essential to TCP performance, that the RTO is being determined accurately, as a too low RTO could prompt unnecessary retransmissions and too high RTO could slow down the application TCP is serving \cite{huston2000tcp,jacobson1995congestion}. To implicitly notify TCP that congestion occurs, congested network nodes will occasionally drop packets after a certain threshold of queued packets has been exceeded. Usually, this is quite smaller than the maximum queue length, therefore avoiding actions can be taken in time \cite{ramakrishnan1999proposal}.\\
Note however, a high RTT can also have other causes than packet loss or congestion. A short term drastic decrease in the receivers processing performance can lead to packets being buffered at the receivers end for a prolonged period of time, thus resulting in a higher RTT \cite{huston2000tcp}.

@@

\section{Network Assisted Congestion Control}
The problem that comes with classic TCP is that the congestion detection algorithms are very implicit. As already mentioned, packet loss for example may have different causes other than network congestion, however, the congestion avoidance algorithm will still interpret it as congestion, thus possibly resulting in unnecessary network delay. Therefore, exactly determining whether the network is congested or not is crucial to keep a network stable. Here, the network layer comes into play. Because most communication should be hidden along the path of the packet, all network nodes will only communicate on the network layer (where IP lives), but not above (where TCP lives) \cite{ramakrishnan1999proposal}.\\
In 2001 a new method was standardized to explicitly notify TCP on congestion. \textit{Explicit Congestion Notification} (ECN) is an extension to TCP and IP, allowing end-to-end congestion notification, while also minimizing the amount of packets dropped. This is done by modifying the IP header to be able to carry one more bit - the \textit{Congestion Experienced} flag (CE) \cite{ramakrishnan1999proposal}. Instead of dropping packets occasionally after the amount of queued packets exceed a threshold [\nameref{P_C_marker}], all packets supporting ECN will now be marked (setting the CE bit in the IP header) [Figure \ref{fig:ecn}]. Thus, congestion can be described explicitly resulting in better short- and long-term network performance. Finally, if a data packet has been marked, the according acknowledgment packet will also be marked to inform the sender of congestion \cite{ramakrishnan1999proposal,ramakrishnan2001addition,10.1145/205511.205512}.\\
To sum up so far, without ECN, classic TCP would just drive the network into congestion, and then recover from it. The network generally speaking will at least experience some jitter (fluctuating network delay). With ECN however, packets in congested nodes will be marked, explicitly informing the TCP sender that the network is congested. As a result, avoiding actions can be taken immediately. In addition, the need to retransmit packets would decrease drastically, as no packets are being dropped anymore to indicate congestion. Therefore, this central algorithm is also being called \textit{Congestion Control} \cite{ramakrishnan1999proposal,ramakrishnan2001addition}.

@@

\subsection{Evaluation of Network Assisted Congestion Control}
As you can see, including the network layer to detect congestion can have great performance improvements. 'Understanding XCP: Equilibrium and Fairness' has shown through simulations that many network topologies could benefit from XCP as it clears queues in equilibrium, thus resulting in improved network performance under load \cite{1498331}. Another paper has shown that XCP outperforms ordinary TCP in nearly any environment related to average queue lengths, efficiency, and throughput \cite{katabi2002congestion}.
\\However, all this performance comes at a cost. Since network nodes have to actively detect congestion, ECN, XCP, and other network assisted congestion control protocols rely on active queue management in those nodes. Although there is hardware being produced today that meet those requirements, a large part of internet hardware isn't capable of doing so. Because any of those nodes can be congested, it would be counterintuitive to run those protocols, as congestion might or might not be detected, thus resulting in performance loss \cite{katabi2002congestion}. This is the reason you do not see those protocols in wide area networks (WAN) too often, let alone the internet.
\\Nevertheless, whenever the implementation of XCP, ECN, or similar protocols is not a challenge, you can implement network assisted congestion control with ease and see great performance improvements. Especially smaller networks with high traffic could definitely benefit from those technologies.

>>>>

\section{Congestion Avoidance}
\label{sec:ca}
As already mentioned, congestion detection algorithms make up an important part in congestion avoidance protocols. \textit{TCP} is such a protocol, based on the transport layer, which adapts its send rate according to available network bandwidth \cite{1209197,jacobson1992tcp}. To fully understand, how TCP's congestion avoidance algorithm operates, one needs to understand how TCP works. TCP lives on the transport layer. It communicates with the application layer via sockets, splits up data into packets (and vice versa), and transmits those packages over the network with the help of the internet protocol, thus TCP is often also referred to as \textit{TCP/IP}. It is the interface between network components and computer programs. Each of those packets have different purposes, but underlay the same structure - the protocol header. The TCP header is split up into multiple segments [Figure \ref{fig:TCP_Header}]:
\begin{figure}
\centering
\begin{bytefield}[bitheight=2.2\baselineskip ,bitwidth=0.635\baselineskip]{32}
	\bitbox{16}{Source Port} & \bitbox{16}{Destination Port} \\
	\bitbox{32}{Sequence Number} \\
	\bitbox{32}{Acknowledgment Number} \\
	\bitbox{4}{Data Offset} & \bitbox{4}{Rsrvd} &
\bitbox{1}{\tiny C\\W\\R} & \bitbox{1}{\tiny E\\C\\E} &
\bitbox{1}{\tiny U\\R\\G} & \bitbox{1}{\tiny A\\C\\K} &
\bitbox{1}{\tiny P\\S\\H} & \bitbox{1}{\tiny R\\S\\T} &
\bitbox{1}{\tiny S\\Y\\N} & \bitbox{1}{\tiny F\\I\\N} &
\bitbox{16}{Window Size} \\
\bitbox{16}{Checksum} & \bitbox{16}{Urgent Pointer} \\
\bitbox{32}{Options} \\
\bitbox{32}{Data}
\end{bytefield}
\caption{Basic structure of the TCP header. It is split up into multiple 32-bit blocks. The data block contains all the application data to be transmitted \cite{jacobson1992tcp,ietf-tcpm-rfc793bis-16,huston2000tcp}.}
\label{fig:TCP_Header}
\end{figure}
\begin{itemize}
\item \textit{Source Port} and \textit{Destination Port} specify the ports on which TCP should communicate with the application layer.

@@

\item TCP can set eight different bit-flags in its header. The most important one is the \textit{ACK} flag (Acknowledgement bit). Whenever a data packet arrives at its destination, a packet acknowledging the arrival of this packet will be send back. This acknowledgement packet always has the \textit{ACK} flag set. The \textit{SYN} and \textit{FIN} flags initialize and terminate a connection.

\item The \textit{Window Size} is the maximum amount of data (in bytes) a receiver is willing to receive at any time. It is important to set it appropriately to prevent unnecessary packet loss or delay, as the receiver couldn't process the data in time. Thus, it is will also be constantly adjusted to improve the connections reliability \cite{jacobson1992tcp}.
\end{itemize}
TCP uses a handshake system. Whenever a client sends a data packet to the server, the server waits until it has received enough data (specified by the window size), and then sends back an acknowledgment package, signaling the successful transmission of data. There are also other handshake operations used, for example for establishing or terminating a connection, however, those have no further importance to the congestion avoidance protocol \cite{1209197,jacobson1992tcp,huston2000tcp,jacobson1995congestion}.

\subsection{Perceiving Congestion}
\label{P_C_marker}
Both partners of a connection are both sender and receiver, as both transmit packets. In the following however, we will simplify this and specify one partner as the sender, the other as the receiver. The sender keeps track of the time it takes from when a package sent until the acknowledgment is received. This is also known as \textit{round-trip time} (RTT). Comparing the current RTT to previously measured RTTs can inform the sender about the current state of congestion. Some variances of TCP make use of this feature in their congestion avoidance algorithms, yet classic TCP does not \cite{huston2000tcp,jacobson1995congestion}. RTT also plays a very important role in the detection of packet loss, which further implies congestion. The sender side keeps track of the RTT while it is being measured and compares it to the \textit{retransmission timeout} (RTO) \cite{jacobson1992tcp}. RTO time is being calculated by estimating the mean and variance of the RTT. Whenever packets are not acknowledged within this RTO interval, the TCP sender regards this packet as lost and will retransmit its data. Therefore, it is essential to TCP performance, that the RTO is being determined accurately, as a too low RTO could prompt unnecessary retransmissions and too high RTO could slow down the application TCP is serving \cite{huston2000tcp,jacobson1995congestion}. To implicitly notify TCP that congestion occurs, congested network nodes will occasionally drop packets after a certain threshold of queued packets has been exceeded. Usually, this is quite smaller than the maximum queue length, therefore avoiding actions can be taken in time \cite{ramakrishnan1999proposal}.

@@

\section{Network Assisted Congestion Control}
The problem that comes with classic TCP is that the congestion detection algorithms are very implicit. Packet loss for example may have different causes other than network congestion, such as simple cable interference, however, the congestion avoidance algorithm will still interpret it as congestion, thus possibly resulting in unnecessary network delay. Therefore, exactly determining whether the network is congested or not is crucial to keep a network stable. Here, the network layer comes into play. Because most communication should be hidden along the path of the packet, all network nodes will only communicate on the network layer (where IP lives), but not above (where TCP lives) \cite{ramakrishnan1999proposal}.\\
In 2001 a new method was standardized to explicitly notify TCP on congestion. \textit{Explicit Congestion Notification} (ECN) is an extension to TCP and IP, allowing end-to-end congestion notification, while also minimizing the amount of packets dropped. This is done by modifying the IP header to be able to carry one more bit - the \textit{Congestion Experienced} flag (CE) \cite{ramakrishnan1999proposal}. Instead of dropping packets occasionally after the amount of queued packets exceed a threshold [\nameref{P_C_marker}], all packets supporting ECN will now be marked (setting the CE bit in the IP header) [Figure \ref{fig:ecn}]. Thus, congestion can be described explicitly resulting in better short- and long-term network performance. Finally, if a data packet has been marked, the according acknowledgment packet will also be marked to inform the sender of congestion \cite{ramakrishnan1999proposal,ramakrishnan2001addition,10.1145/205511.205512}.\\
To sum up so far, without ECN, classic TCP would just drive the network into congestion, and then recover from it. The network generally speaking will at least experience some jitter (fluctuating network delay). With ECN however, packets in congested nodes will be marked, explicitly informing the TCP sender that the network is congested. As a result, avoiding actions can be taken immediately. In addition, the need to retransmit packets would decrease drastically, as no packets are being dropped anymore to indicate congestion. Therefore, this central algorithm is also being called \textit{Congestion Control} \cite{ramakrishnan1999proposal,ramakrishnan2001addition}.

@@

\subsection{Evaluation of Network Assisted Congestion Control}
As you can see, including the network layer to detect congestion can have great performance improvements. 'Understanding XCP: Equilibrium and Fairness' has shown through simulations that many network topologies could benefit from XCP as it clears queues in equilibrium, thus resulting in improved network performance under load \cite{1498331}. Another paper has shown that XCP outperforms ordinary TCP in nearly any environment related to average queue lengths, efficiency, and throughput \cite{katabi2002congestion}.
\\However, all this performance comes at a cost. Since network nodes have to actively detect congestion, ECN, XCP, and other network assisted congestion control protocols rely on active queue management in those nodes. Although there is some hardware being produced today that meet ECNs requirements, a large part of internet hardware isn't capable of doing so. Because any of those nodes can be congested, it would be counterintuitive to run those protocols, as congestion might or might not be detected, thus resulting in performance loss \cite{katabi2002congestion}. This is the reason you do not see those protocols in wide area networks (WAN) too often, let alone the internet.
\\Nevertheless, whenever the implementation of XCP, ECN, or similar protocols is not a challenge, you can implement network assisted congestion control with ease and see great performance improvements. Especially smaller networks with high traffic could definitely benefit from those technologies.
